---
title: "Myeloid cells with SUSD1"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction

## Objective
In this Rmarkdown document, we are going to 

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(gt)
library(grid)
library(RColorBrewer)
library(parallelDist)
library(scales)
library(gtable)
library(ggpubr)
library(rstatix)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/CLARIAJOA_metadata.csv")
path_project_clinical <- here::here("01_cellranger_mapping/data/CLARIAJOA_samplesdata_clinical.xlsx")
path_r_objects <- here::here("04_clustering_annotation/results/R_objects")
path_r_figs_out <- here::here("ACLF_figures/figs")

# Functions
source(here::here("bin/utils.R"))

# Parameters
confounder_variables <- c("library_name", "hashing", "patient", "treatment", "disease_grade", "gender")
confounder_names <- c("Library", "Hashing", "Patient", "Treatment", "Disease", "Gender")

qc_metrics <-  c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb", "S.Score", "G2M.Score")

palette_color_cells <- c("Classical Monocytes" = "#6b4e71",
                         "Intermediate Monocytes HAVCR2+" = "#8ecae6",
                         "Intermediate Monocytes IFN+" = "#023047",
                         "Intermediate Monocytes Inflammatory" = "#4c5760",
                         "Intermediate Monocytes Activated" = "#40798c",
                         "Non-Classical Monocytes" = "#E56B6F",
                         "Macrophages-like Monocytes" = "#e4c1f9",
                         "DC2" = "#ffb703",
                         "DC4 FCGR3B+" = "#fb8500",
                         "pDCs" = "#ef233c",
                         "G-MDSC" = "#cfe1b9",
                         "G-MDSC CST7+" = "#3f7d20",
                         "GMP cells" = "#d4a373"
)

palette_color_treatment <- c("HSA" = "#ee6c4d",
                             "vehicle" = "#3d5a80")

palette_color_disease <- c("AD" = "#ffe1a8",
                               "ACLFg1" = "#E26D5C",
                               "ACLFg2" = "#822121",
                               "ACLFg3" = "#723d46",
                               "Healthy" = "#99EDC3")

palette_color_renalfunction <- c("Healthy" = "#3DED97",
                                 "No Renal D/F" = "#FF8A8A",
                                 "Renal D/F" = "#9B111E")
```

## Load metadata
The data used in this Rmarkdown document comes from **CLARIAJOA** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata_all tables
metadata <- metadata %>% filter(type == "cDNA")

print("Libraries metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))

# Clinical metadata
clinical_metadata <- openxlsx::read.xlsx(path_project_clinical)
print("Clinical metadata")
DT::datatable(clinical_metadata, options = list(scrollX = TRUE))
```

## Load data
```{r}
# Load Seurat object
seurat_obj <- readRDS(here::here("04_clustering_annotation/results/R_objects/CLARIAJOA_myeloidcells_clustering_level3_annotation_cleaned_subset.rds"))
seurat_obj

# Subsetting data only PATIENTS, removing HC
Idents(seurat_obj) <- "disease"
Idents(seurat_obj) <- "annot_level3"

# Reorder factors
seurat_obj$annot_level3 <- factor(x = seurat_obj$annot_level3,
                                    levels = c("Classical Monocytes",
                                               "Intermediate Monocytes Activated",
                                               "Intermediate Monocytes HAVCR2+",
                                               "Intermediate Monocytes Inflammatory",
                                               "Intermediate Monocytes IFN+",
                                               "Non-Classical Monocytes",
                                               "Macrophages-like Monocytes",
                                               "DC2",
                                               "DC4 FCGR3B+",
                                               "pDCs",     
                                               "G-MDSC",
                                               "G-MDSC CST7+",
                                               "GMP cells"
                                               ))

# Merge clinical data with seurat object
clinical_metadata <- clinical_metadata[, !colnames(clinical_metadata) %in% c("ID", "ACLF.grade", "ACLF_yesno")]

seurat_obj_metadata <- left_join(seurat_obj@meta.data, clinical_metadata, by = "patient")
rownames(seurat_obj_metadata) <- rownames(seurat_obj@meta.data)
seurat_obj@meta.data <- seurat_obj_metadata


# Subset untreated
seurat_obj <- subset(seurat_obj, subset = treatment == "vehicle")

# New variable
seurat_obj$disease_general <- as.character(seurat_obj$disease)
seurat_obj$disease_general[seurat_obj$disease == "AD"] <- "Decompensation"
seurat_obj$disease_general[seurat_obj$disease == "ACLF"] <- "Decompensation"
seurat_obj$disease_general <- as.factor(seurat_obj$disease_general)

# Subset only patients
seurat_obj_sub <- subset(seurat_obj, subset = disease %in% c("AD", "ACLF"))
```


```{r}
table(seurat_obj$annot_level3, seurat_obj$disease)
```


## SUSD1
```{r fig.width=8, fig.height=7}
DotPlot(seurat_obj, features = "SUSD1", group.by = "annot_level3", split.by = "disease_general", cols = "RdBu")
```

```{r fig.width=12, fig.height=5}
VlnPlot(seurat_obj, features = "SUSD1", group.by = "annot_level3", split.by = "disease_general") &
  stat_compare_means(method = "wilcox", ref.group = "Healthy")
```


## SUSD1 by clinical condition
```{r}
clinical_conditions <- c("Liver.subscore", 
                         "Kidney.subscore",
                         "Coagulation.subscore",
                         "Circulation.subscore",
                         "Lung.subscore", 
                         "Cerebral.subscore")


# Replace missing values with "Healthy" for each variable column with missing information
for (col in clinical_conditions) {
  seurat_obj@meta.data[[col]][is.na(seurat_obj@meta.data[[col]])] <- "Healthy"
}
```


```{r fig.width=12, fig.height=5}
pdf("figs/CLARIAJOA_Myeloid_SUSD1_byclinicalcond.pdf", width = 12, height = 5, onefile = TRUE)
for (cond in clinical_conditions) {
  gg <- VlnPlot(seurat_obj, 
        features = "SUSD1", 
        group.by = cond, 
        split.by = "annot_level3") +
    scale_fill_manual(values = palette_color_cells) +
    labs(x = cond)
  print(gg)
}

dev.off()
```


```{r fig.width=12, fig.height=5}
pdf("figs/CLARIAJOA_Myeloid_patients_SUSD1_byclinicalcond.pdf", width = 12, height = 5, onefile = TRUE)
for (cond in clinical_conditions) {
  gg <- VlnPlot(seurat_obj_sub, 
        features = "SUSD1", 
        group.by = cond, 
        split.by = "annot_level3") +
    scale_fill_manual(values = palette_color_cells) +
    labs(x = cond)
  print(gg)
}
dev.off()
```



## SUSD1 by manual kidney_function condition
```{r}
seurat_obj$kidney_function <- ifelse(seurat_obj$`Creat(mg/dl)` >= 2, "Renal failure",
                                     ifelse((seurat_obj$`Creat(mg/dl)` < 2 & seurat_obj$`Creat(mg/dl)` >= 1.5 ), "Renal dysfunction", 
                                            "No Renal dysfunction/failure"))
                                     
seurat_obj$kidney_function[seurat_obj$disease == "Healthy"] <- "Healthy"    

table(seurat_obj$kidney_function)
```


```{r fig.width=8, fig.height=9}
DotPlot(seurat_obj, features = "SUSD1", group.by = "annot_level3", split.by = "kidney_function", cols = "RdBu")
DotPlot(seurat_obj, features = "SUSD1", group.by = "annot_level3", split.by = "kidney_function", cols = "RdBu", col.max = 1)
```

```{r fig.width=12, fig.height=5}
# Create new variable based on cluster level
pdf("figs/CLARIAJOA_Myeloid_SUSD1_byCreatinine.pdf", width = 12, height = 5, onefile = TRUE)

VlnPlot(seurat_obj, 
        features = "SUSD1", 
        group.by = "kidney_function", 
        split.by = "annot_level3") +
    scale_fill_manual(values = palette_color_cells) +
    labs(x = cond)

dev.off()
```


```{r fig.width=12, fig.height=5}
seurat_obj_sub$kidney_function <- ifelse(seurat_obj_sub$`Creat(mg/dl)` >= 2, "Renal failure",
                                     ifelse((seurat_obj_sub$`Creat(mg/dl)` < 2 & seurat_obj_sub$`Creat(mg/dl)` >= 1.5 ), "Renal dysfunction", 
                                            "No Renal dysfunction/failure"))
table(seurat_obj_sub$kidney_function)
```


```{r fig.width=12, fig.height=5}
# Create new variable based on cluster level
pdf("figs/CLARIAJOA_Myeloid_patients_SUSD1_byCreatinine.pdf", width = 12, height = 5, onefile = TRUE)

VlnPlot(seurat_obj_sub, 
        features = "SUSD1", 
        group.by = "kidney_function", 
        split.by = "annot_level3") +
    scale_fill_manual(values = palette_color_cells) +
    labs(x = cond)

dev.off()
```


## SUSD1 by manual kidney_function condition simplified
```{r}
seurat_obj$kidney_function_simple <- ifelse(seurat_obj$`Creat(mg/dl)` >= 1.5, "Renal D/F", "No Renal D/F")
seurat_obj$kidney_function_simple[seurat_obj$disease == "Healthy"] <- "Healthy"    

# Remove patient without information -- missing information included
# seurat_obj <- subset(seurat_obj, subset = donor_id == "ACLF8", invert = TRUE)

# Reorder factors
seurat_obj$kidney_function_simple <-  factor(x = seurat_obj$kidney_function_simple,
                                    levels = c("Healthy",
                                               "No Renal D/F",
                                               "Renal D/F"))
table(seurat_obj$kidney_function_simple)
table(seurat_obj$annot_level3, seurat_obj$kidney_function_simple)
```


```{r fig.width=8, fig.height=7}
DotPlot(seurat_obj, features = "SUSD1", group.by = "annot_level3", split.by = "kidney_function_simple", cols = "RdBu")
DotPlot(seurat_obj, features = "SUSD1", group.by = "annot_level3", split.by = "kidney_function_simple", cols = "RdBu", col.max = 1)
```

```{r fig.width=20, fig.height=6}
# Create new variable based on cluster level
VlnPlot(seurat_obj, 
        features = "SUSD1", 
        group.by = "annot_level3", 
        split.by = "kidney_function_simple") &
    scale_fill_manual(values = palette_color_renalfunction)
```


### Manual plotting for statistics
```{r}
vln_df = data.frame(SUSD1 = seurat_obj[["RNA"]]@data["SUSD1", ], 
                    annot_level3 = seurat_obj$annot_level3,
                    kidney_function_simple = seurat_obj$kidney_function_simple)
#noise <- rnorm(n = length(x = vln_df[, "SUSD1"])) / 100000
#vln_df$SUSD1_noise <- vln_df$SUSD1 + noise

# Reorder factors
vln_df$kidney_function_simple <-  factor(x = vln_df$kidney_function_simple,
                                    levels = c("Healthy",
                                               "No Renal D/F",
                                               "Renal D/F"))
vln_df$SUSD1 <- as.numeric(vln_df$SUSD1)
```


```{r}
mycomparisons <- list(c("No Renal D/F", "Healthy"),
                      c("Renal D/F", "No Renal D/F"),
                      c("Renal D/F", "Healthy")
                      )

# The Wilcoxon test requires at least 3 observations in each group in order to be performed.
table(vln_df$annot_level3, vln_df$kidney_function_simple)
vln_df_count <- vln_df %>% 
  group_by(annot_level3, kidney_function_simple) %>% 
  summarise(diff_0 = sum(SUSD1 != 0))
print(vln_df_count)
# As seen, G-MDSC / G-MDSC CST7+ / GMP cells population does not fulfill, will be excluded.

# Filtering populations
vln_df <- vln_df %>% 
  filter(!annot_level3 %in% c("G-MDSC", "G-MDSC CST7+", "GMP cells"))

# Computing significance
stat.test <- vln_df %>%
  group_by(annot_level3) %>%
  wilcox_test(SUSD1 ~ kidney_function_simple, comparisons = mycomparisons, alternative = "two.sided") %>%
  adjust_pvalue(method = "fdr") %>%
  add_significance()

write.table(as.data.frame(stat.test),
          file = "figs/CLARIAJOA_Myeloid_SUSD1_byCreatinine_simplified_statistics.csv",
          sep = ",", row.names = FALSE)

stat.test <- stat.test %>% add_xy_position()
stat.test
```



```{r fig.width=20, fig.height=8}
pdf("figs/CLARIAJOA_Myeloid_SUSD1_byCreatinine_simplified.pdf", width = 20, height = 8, onefile = TRUE)

#  violin plot without noise
vln_df %>% 
  ggplot(aes(x = kidney_function_simple, y = SUSD1, fill = kidney_function_simple)) + 
  geom_violin(adjust =1, trim=TRUE, scale = "width") +
  facet_wrap(annot_level3 ~ ., scales = "free", ncol = 5) +
  scale_fill_manual(values = palette_color_renalfunction) +
  geom_jitter(alpha=0.2) +
  #stat_pvalue_manual(stat.test, tip.length = 0.01, label = "{scales::pvalue(p.adj)}") +
  #stat_compare_means(comparisons = mycomparisons, method = "wilcox.test", label = "p.format") +
  labs(x = "", y = "SUSD1 Expression Levels", fill = "Patient stratification") +
  theme(strip.text = element_text(size = 15),
        axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.text = element_text(size=14),
        legend.position = "bottom") +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.15)))

dev.off()
```



# Session Info
```{r}
sessionInfo()
```