---
title: "Annotation: Level 1"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
After the pre-processing steps done in previous notebooks, we will now start with the downstream data analysis in order to extract biological insights and describe the underlying biology.

## Objective
In this Rmarkdown document, we are going to assign cell type clusters into different cell-lineage clusters based on the similarity of their gene expression profiles. Then, we will perform again process each object to define cell-type specific clusters with more resolution.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(DT)
library(dittoSeq)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/CLARIAJOA_metadata.csv")

path_r_objects <- here::here("04_clustering_annotation/results/R_objects")
path_r_figs <- here::here("04_clustering_annotation/results/figs")
path_r_tables <- here::here("04_clustering_annotation/results/tables")

# Functions
source(here::here("bin/utils.R"))

# Parameters
qc_metrics <-  c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb", "scrublet_doublet_scores")

confounder_variables <- c("library_name", "hashing", "patient", "treatment", "disease_grade", "gender")
confounder_names <- c("Library", "Hashing", "Patient", "Treatment", "Disease", "Gender")

# Color palette
set.seed(1234) # for reproducibility
color_palette <- Polychrome::createPalette(32, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```

## Load data
The data used in this Rmarkdown document comes from **CLARIAJOA** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata_all tables
metadata <- metadata %>% filter(type == "cDNA")

print("Libraries metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))

## Define resolution of choice
resolution <- 0.05
resolution_name <- paste0("RNA_snn_res.", resolution)

# Load Seurat object
seurat_obj <- readRDS(paste0(path_r_objects, "/CLARIAJOA_clustering_level1_resolutions.rds"))
DefaultAssay(seurat_obj) <- "RNA"
Idents(seurat_obj) <- "RNA_snn_res.0.05"
seurat_obj

# Add variable
seurat_obj$disease_grade_treatment <- paste0(seurat_obj$disease_grade, "_", seurat_obj$treatment)
seurat_obj$disease_grade_treatment_patient <- paste0(seurat_obj$disease_grade, "_", seurat_obj$treatment, "_", seurat_obj$patient)

# Load biomarkers
seurat_obj_markers <- readRDS(paste0(path_r_objects, "/CLARIAJOA_biomarkers_level1_res0.05.rds"))
```


# Explore cell clusters
To visualize the cell clusters, there are a few different dimensionality reduction techniques, such as t-SNE and UMAP. These methods require to input a number of PCA dimensions (same number as the one used for the clustering). We will not run UMAP again as it was computed in the previous notebook.
```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering <- DimPlot(object = seurat_obj,
        group.by = resolution_name,
        label = TRUE,
        label.size = 5,
        cols = color_palette,
        raster = FALSE,
        ) & NoLegend()
gg_umap_clustering
```

Now, we will also explore clusters across therapy.
```{r fig.width=14, fig.height=6}
# Clusters across therapys
gg_clustering_bytherapy_umap <- DimPlot(
  object = seurat_obj,
  group.by = resolution_name,
  split.by = "disease_grade",
  label = FALSE,
  cols = color_palette
  )

gg_clustering_bytherapy_umap
```


## Cell clusters data exploration
```{r}
# Number of cells / cluster
table(seurat_obj@active.ident)

# Number of cells per cluster vs confounder variables
table(seurat_obj$library_name, seurat_obj@active.ident)
table(seurat_obj$disease_grade, seurat_obj@active.ident)
table(seurat_obj$treatment, seurat_obj@active.ident)
```


## Assessing sources of uninteresting variation
Next, we will explore additional metrics, such as the number of UMIs and genes per cell, mitochondrial / ribosomal gene expression and scrublet doublet scores on the UMAP embedding. 

```{r fig.width=14, fig.height=15}
gg_qc_umap <- FeaturePlot(
  seurat_obj,
  reduction = "umap",
  features = qc_metrics,
  cols = c("lightgrey", "darkblue"), 
  order = TRUE,
  ncol = 2)

gg_qc_umap
```

```{r fig.width=10, fig.height=15}
gg_qc_vln <- VlnPlot(
  seurat_obj,
  features = qc_metrics,
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_qc_vln
```

Finally, we will also check the cell-cycle phase by assessing S-phase and G2M-phase markers signature.
```{r fig.width=10, fig.height=10}
gg_phase_vln <- VlnPlot(
  seurat_obj,
  features = c("S.Score", "G2M.Score"),
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_phase_vln
```

# Explore cluster biomarkers
## Table
```{r}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

top_all_markers
#DT::datatable(top_all_markers)
```

## Dotplot
```{r fig.width=16, fig.height=6}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 4, order_by = avg_log2FC)

top_genemarkers_list <- unique(top_all_markers$gene)

DotPlot(
  seurat_obj,
  features = top_genemarkers_list,
  dot.scale = 6) +
  RotatedAxis()
```


<!-- ## FeaturePlot and ViolinPlot -->
<!-- With the cells clustered and based on the previous cell-type markers, we will select the most informative gene markers for each cluster and some well-known cell markers and project them in the UMAP. Basically, we are looking for consistency of expression of the markers across the clusters (e.g. if there are two markers for a cell type and only one of them is expressed in a cluster, we cannot reliably assign that cluster to the cell type). -->
<!-- ```{r} -->
<!-- # Selected gene markers -->
<!-- selected_markers <- seurat_obj_markers %>% -->
<!--   group_by(cluster) %>% -->
<!--   slice_max(n = 4, order_by = avg_log2FC) %>% -->
<!--   group_split() -->
<!-- ``` -->

<!-- ```{r fig.width=7, fig.height=5} -->
<!-- # Feature plots -->
<!-- purrr::map(selected_markers, function(cluster) { -->
<!--   FeaturePlot(seurat_obj, -->
<!--               features = cluster$gene, -->
<!--               label = TRUE, -->
<!--               label.size = 2, -->
<!--               ncol = 2) -->
<!-- }) -->
<!-- ``` -->

## Explore gene markers

```{r fig.width=25, fig.height=20}
lineage_genes <- c("PTPRC", # CD45+, immune cells
                   "CD3D", "CD3E", # T cells
                   "CD4",  "IRF7", "FOXP3", # CD4+ T cells, Treg
                   "CD8A", "CD8B", "NKG7", # CD8+ T cells, cytotoxic
                   "KLRD1", "GZMA", # NK cells
                   "CD79A", # B cells
                   "MS4A1", "JCHAIN", # Naive vs Plasma cells
                   "LYZ", "CD14", "FCGR3A", "PLAC8", "CD209", # Mono/macro/DCs
                   "S100A9", "CD200R3", # Neutrophils, Mast cells
                   "MKI67", "TOP2A", "HLA-DRA", # Cycling, MHC-antigen presenting
                   "PPBP"
                   )

FeaturePlot(seurat_obj,
            features = lineage_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5)

VlnPlot(
  seurat_obj,
  features = lineage_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```



# Annotation
According to the previous cell-type identities enriched in several gene markers, we will establish the following annotation:

```{r}
seurat_obj <- RenameIdents(
  seurat_obj,
 "0" = "T-NK cells",
 "1" = "Myeloid cells",
 "2" = "Myeloid cells",
 "3" = "B Lymphocytes", # B cells
 "4" = "Mixed cells", # Plasma cells & proliferating cells
 "5" = "Granulocytes",  # B cells
 "6" = "Platelet complexes",
 "7" = "Myeloid cells", # pDCs
 "8" = "Low quality cells"
)

# Creating new annotation variable
seurat_obj$annot_level1 <- as.character(seurat_obj@active.ident)
Idents(seurat_obj) <- "annot_level1"

# Exploring number of celltypes
table(seurat_obj$annot_level1)
```


## UMAP celltypes
```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering_annot <- DimPlot(object = seurat_obj,
        group.by = "annot_level1",
        label = TRUE,
        label.size = 5,
        cols = color_palette
        ) +
  labs(title = "Level 1")
gg_umap_clustering_annot

ggsave(filename = paste0(path_r_figs, "/CLARIAJOA_clustering_level1_annotation_umap.png"),
       plot = gg_umap_clustering_annot)
```


Now, we will also explore clusters across therapy
```{r fig.width=15, fig.height=5}
DimPlot(
  object = seurat_obj,
  group.by = "annot_level1",
  split.by = "disease_grade",
  raster = FALSE,
  label = FALSE,
  cols = color_palette
  ) + NoLegend()
```



## Cell proportions
```{r fig.width=14, fig.height=6}
gg_cell_num <- dittoBarPlot(seurat_obj, 
                            "annot_level1",
                            group.by = "disease_grade_treatment_patient",
                            scale = "count",
                            legend.show = FALSE,
                            #retain.factor.levels = TRUE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj, 
                             "annot_level1",
                             group.by = "disease_grade_treatment_patient",
                             scale = "percent",
                            #retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```


```{r fig.width=15, fig.height=6}
gg_cell_num <- dittoBarPlot(seurat_obj, 
                            "annot_level1",
                            group.by = "disease_grade_treatment",
                            scale = "count",
                            legend.show = FALSE,
                            #retain.factor.levels = TRUE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj, 
                             "annot_level1",
                             group.by = "disease_grade_treatment",
                             scale = "percent",
                            #retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```

## Save Seurat object
```{r}
# Save lineage annotations
saveRDS(seurat_obj$annot_level1, paste0(path_r_objects, "/CLARIAJOA_annotation_level1.rds"))

# Save processed lognorm Seurat object
saveRDS(seurat_obj, paste0(path_r_objects, "/CLARIAJOA_clustering_level1_annotation.rds"))
#seurat_obj <- readRDS(paste0(path_r_objects, "/CLARIAJOA_clustering_level1_annotation.rds"))
```


# Generate Seurat objects (by cell lineage)
## Split object by cell lineage
```{r}
# Preserving only metadata of interest
seurat_metadata_variables <- c(names(seurat_obj@meta.data[1:38]), "annot_level1")
seurat_obj@meta.data <- seurat_obj@meta.data[seurat_metadata_variables]
```

## Save Seurat objects
```{r}
# Mono/macro/DCs
seurat_obj_myeloid <- subset(seurat_obj, idents = "Myeloid cells")
saveRDS(seurat_obj_myeloid, paste0(path_r_objects, "/CLARIAJOA_clustering_level1_myeloidcells.rds"))
rm(seurat_obj_myeloid)

# B Lymphocytes
seurat_obj_blymphocytes<- subset(seurat_obj, idents = "B Lymphocytes")
saveRDS(seurat_obj_blymphocytes, paste0(path_r_objects, "/CLARIAJOA_clustering_level1_Blymphocytes.rds"))
rm(seurat_obj_blymphocytes)

# T + NK cells
seurat_obj_TNKcells <- subset(seurat_obj, idents = "T-NK cells")
saveRDS(seurat_obj_TNKcells, paste0(path_r_objects, "/CLARIAJOA_clustering_level1_TNKcells.rds"))
rm(seurat_obj_TNKcells)

# Mixed cells
seurat_obj_other <- subset(seurat_obj, idents = "Mixed cells")
saveRDS(seurat_obj_other, paste0(path_r_objects, "/CLARIAJOA_clustering_level1_mixedcells.rds"))
rm(seurat_obj_other)

# Not subclutering: Low quality cells (will be removed), platelet complexes & granulocytes!
```


# Session Info
```{r}
sessionInfo()
```
