---
title: "Quality control doublet prediction not-hashed libraries"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
## Objective
In this Rmarkdown document, we are going to summarize the results obtained by the `scrublet` doublet detection step to flag possible "Doublets" present in the not-hashed data, for which there is no background truth for barcodes containing more than one cell. Finally, we will merge all not-hashed libraries into a single Seurat object.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/CLARIAJOA_metadata.csv")
path_samples_metadata <- here::here("01_cellranger_mapping/data/CLARIAJOA_metadata_samples_all.csv")
path_save_seuratobj <- here::here("02_demultiplexing/results/R_objects")

# Functions
source(here::here("bin/utils.R"))
```

## Load data
The data used in this Rmarkdown document comes from **CLARIAJOA** dataset, filtered matrices were processed with `cellranger v6.1.1` and the doublet predictions were obtained using `scrublet`.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)
metadata <- metadata[order(metadata$gem_id), !names(metadata) %in% "project", drop = F]

metadata_samples <- read.csv(path_samples_metadata)

# Get list of GEM_IDs for not-hashed samples
not_hashed_gemids <- as.vector(sort(unique(metadata$gem_id[metadata$hashing == "not_hashed"])))

# Merging project and samples metadata tables
metadata <- metadata %>%
  filter(type == c("cDNA")) %>%
  filter(gem_id %in% not_hashed_gemids)
metadata_all <- left_join(metadata, metadata_samples, by = "patient")

print("Libraries metadata")
DT::datatable(metadata_all, options = list(scrollX = TRUE))
```

```{r}
# Create Seurat object list
seurat_obj_not_hashed_list <- purrr::map(not_hashed_gemids, function(gem_id) {
  path_filtered_matrix <- here::here(paste(
  "01_cellranger_mapping/jobs", gem_id, gem_id, "outs/filtered_feature_bc_matrix", sep = "/"))
  counts_matrix <- Read10X(data.dir = path_filtered_matrix)
  seurat = CreateSeuratObject(counts = counts_matrix)
  # Add gem_id to each barcode
  seurat$gem_id <- gem_id
  seurat
})
names(seurat_obj_not_hashed_list) <- not_hashed_gemids


# Add doublet predictions
scrublet_files_list <- purrr::map(not_hashed_gemids, function(gem_id) {
  path_scrublet <- here::here(paste0(
  "02_demultiplexing/results/doublet_prediction/tables/", gem_id, "_doublet_prediction_scrublet.csv"))
})
scrublet_dfs <- lapply(scrublet_files_list, function(scrublet_file_path){
  read.csv(scrublet_file_path)
})
names(scrublet_dfs) <- not_hashed_gemids

# Integrate scrublet doublet predictions to seurat metadata
seurat_obj_not_hashed_list <- purrr::map2(seurat_obj_not_hashed_list, 
                                          names(seurat_obj_not_hashed_list), 
                                          function(seurat, gem_id) {
  if (all(colnames(seurat) == scrublet_dfs[[gem_id]]$barcodes)) {
    seurat$scrublet_doublet_scores <- scrublet_dfs[[gem_id]]$scrublet_doublet_scores
    seurat$scrublet_predicted_doublet <- scrublet_dfs[[gem_id]]$scrublet_predicted_doublet
    seurat
  } 
})

seurat_obj_not_hashed_list
```

# Doublet prediction with Scrublet

```{r fig.width = 10, fig.height = 25}
# Comparison by library
gg_scrublet_list <- purrr::map2(seurat_obj_not_hashed_list, 
                                names(seurat_obj_not_hashed_list), 
                                function(seurat, gem_id) {
                                  
  # Retrieving approx. scrublet doublet predicted threshold
  scrublet_threshold <- round(median(
    c(max(seurat$scrublet_doublet_scores[seurat$scrublet_predicted_doublet == "False"]),
      min(seurat$scrublet_doublet_scores[seurat$scrublet_predicted_doublet == "True"]))), 2)
  
  # Plot scrublet doublet scores & predictions
  gg_scrublet_pred <- gg_nothashed_scrublet_pred(seurat@meta.data, gem_id, scrublet_threshold)
  
  # Plot scrublet doublet predictions & UMIs
  gg_umis <- gg_nothashed_umis(seurat)
  
  gg_scrublet_pred + gg_umis
})

cowplot::plot_grid(plotlist = gg_scrublet_list,
                   nrow = length(seurat_obj_not_hashed_list))
```

# Merge Seurat objects
```{r}
seurat_obj_not_hashed <- merge(seurat_obj_not_hashed_list[[1]], 
                    y = seurat_obj_not_hashed_list[2:length(seurat_obj_not_hashed_list)], 
                    add.cell.ids = not_hashed_gemids)

# Add gem_id to barcode
new_barcodes <- str_c(seurat_obj_not_hashed$gem_id, colnames(seurat_obj_not_hashed), sep = "_")
seurat_obj_not_hashed <- RenameCells(seurat_obj_not_hashed, new.names = new_barcodes)
seurat_obj_not_hashed

# Remove Seurat object list
rm(seurat_obj_not_hashed_list)
```

# Add sample metadata
Now, we will add samples' metadata in the Seurat object. In this project, each library contains samples from a single donor. Therefore, we will merge samples metadata by its donor ID, and considering that all samples that were not multiplex only contain blood, we will also include the tissue of origin variable to match with donors for which different tissues were processed.
```{r}
# Create dataframe with seurat metadata and samples & project
seurat_obj_metadata <- left_join(seurat_obj_not_hashed@meta.data, metadata_all, by = "gem_id")

# Add metadata information to Seurat object
rownames(seurat_obj_metadata) <- rownames(seurat_obj_not_hashed@meta.data)
seurat_obj_not_hashed@meta.data <- seurat_obj_metadata
```

# Save Merged Seurat object
```{r}
saveRDS(seurat_obj_not_hashed, 
        paste0(path_save_seuratobj, "/CLARIAJOA_nothashed_merged.rds"))
```

# Session Info
```{r}
sessionInfo()
```
