---
title: "Data integration and processing"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
Following with the results obtained in the previous notebook, we will now perform batch-effect correction (data integration). To do so, we will follow the current best practices for scRNA-seq QC described in [Luecken al. Mol Syst Biol (2018)](https://doi.org/10.15252/msb.20188746) and adapt some workflows from [Satija Lab](https://satijalab.org/seurat/vignettes.html).

## Objective
In this Rmarkdown document, we are going to perform the previous processing steps.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(harmony)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/CLARIAJOA_metadata.csv")
path_samples_metadata <- here::here("01_cellranger_mapping/data/CLARIAJOA_metadata_samples_all.csv")
path_r_objects <- here::here("03_QC/results/R_objects")

# Functions
source(here::here("bin/utils.R"))

# Parameters
confounder_variables <- c("library_name", "hashing", "patient", "treatment", "disease_grade", "gender")
confounder_names <- c("Library", "Hashing", "Patient", "Treatment", "Disease", "Gender")

qc_metrics <-  c("nCount_RNA", "nFeature_RNA", "S.Score", "G2M.Score", "pct_mt", "pct_rb")

# Color palette
color_palette <- Polychrome::createPalette(25, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```

## Load data
The data used in this Rmarkdown document comes from **CLARIAJOA** project.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)
metadata_samples <- read.csv(path_samples_metadata)

# Merging metadata_all tables
metadata_all <- left_join(metadata, metadata_samples, by = "patient")

print("Libraries metadata")
DT::datatable(metadata_all, options = list(scrollX = TRUE))

# Load Seurat object
seurat_obj <- readRDS(paste0(path_r_objects, "/CLARIAJOA_alldata_lognorm_processed.rds"))
seurat_obj
```

# Integration with harmony
The integration approach was carried out with [Harmony](https://www.nature.com/articles/s41592-019-0619-0). [Chazarra-Gil *et al*](https://www.biorxiv.org/content/10.1101/2020.05.22.111211v2), which was reported to be among the best scRNA-seq integration methods, especially in terms of scalability, in a recent integration benchmarking study.

```{r}
seurat_obj_integrated <- seurat_obj %>%
  RunHarmony(group.by.vars = "gem_id")

# Elbowplot to select relevant PCs to analyse the dataset
ElbowPlot(seurat_obj_integrated, ndims = 50, reduction = "harmony")
```

As seen in the previous notebook, we will use the first 20 PCs to visualize data and perform the clustering.
```{r}
# Dimensionality reduction
seurat_obj_integrated <- seurat_obj_integrated %>%
  RunUMAP(reduction = "harmony", dims = 1:20)
```


## Assessing confounder variables
```{r fig.width=16, fig.height=15}
# Visualizing UMAP
gg_umap_uncorrected <- purrr::map2(confounder_variables, confounder_names,
                                          function(confounder, title) {
  p <- DimPlot(seurat_obj_integrated, 
               reduction = "umap",
               group.by = confounder,
               pt.size = 0.01) +
    labs(title=title) +
    scale_color_manual(values=color_palette)
  p
})

cowplot::plot_grid(plotlist = gg_umap_uncorrected, ncol = 2)
```

Here, from the UMAP mapping, we cannot see the presence of library-specific effects that should not be present. But, we clearly see treatment/disease differences. Thus, we will integrate the different samples to harmonize these cells.

## Assessing cell cycle effects
```{r fig.width=7, fig.height=5}
gg_umap_by_phase <- DimPlot(
  seurat_obj_integrated,
  reduction = "umap",
  group.by = "Phase",
  pt.size = 0.01) +
  labs(color = "Cell Cycle Phase")

gg_umap_by_phase
```
Finally, we ensure that there is no batch effect associated to the cell cycle phase.


## Assessing sources of uninterested variation
Next we will explore additional metrics, such as the number of UMIs and genes per cell, S-phase and G2M-phase markers, and mitochondrial / ribosomal gene expression on the UMAP embedding.

```{r fig.width=14, fig.height=15}
gg_qc_umap <- FeaturePlot(
  seurat_obj_integrated,
  reduction = "umap",
  features = qc_metrics,
  order = TRUE,
  ncol = 2)

gg_qc_umap
```


Finally, we'll also check for the doublet score and doublet prediction obtained using `scrublet`.
```{r fig.width=14, fig.height=5}
gg_doubletscore_umap <- FeaturePlot(
  seurat_obj_integrated,
  reduction = "umap",
  features = "scrublet_doublet_scores",
    cols = c("lightgray", "darkblue"),
  order = TRUE)

gg_doubletpred_umap <- DimPlot(
  seurat_obj_integrated,
  reduction = "umap",
  group.by = "scrublet_predicted_doublet")

cowplot::plot_grid(gg_doubletscore_umap,
                   gg_doubletpred_umap, ncol = 2)
```


## Assessing multiple confounder variables
```{r fig.width=14, fig.height=5}
DimPlot(
  seurat_obj_integrated,
  group.by = "disease_grade",
  split.by = "treatment",
  pt.size = 0.01
  ) +
  labs(title = "Treatment",
       color = "Disease") +
  scale_color_brewer(palette = "Paired")
```

```{r fig.width=25, fig.height=5}
DimPlot(
  seurat_obj_integrated,
  group.by = "treatment",
  split.by = "disease_grade",
  pt.size = 0.01
  ) +
  labs(title = "Disease",
       color = "Treatment") +
  scale_color_brewer(palette = "Paired")
```


## Assessing gene markers expression
```{r fig.width=30, fig.height=15}
# Immune fraction
FeaturePlot(seurat_obj_integrated,
            features = c("PTPRC", # CD45+, immune cells
                         "CD3D", # T cells
                         "CD4", "FOXP3", "SELL", "IRF7", "KLRG1", # CD4+ T cells, Treg
                         "CD8A", "NKG7", # CD8+ T cells, cytotoxic
                         "KLRD1", "IL7R", # NK cells
                         "CD79A", # B cells
                         "MS4A1", "JCHAIN", # Naive vs Plasma cells
                         "LYZ", "CD14", "FCGR3A", "PLAC8", # Mono/macro
                         "CLEC9A", "CLEC10A", "FTL", "GZMB", # DCs
                         "S100A9", # Neutrophils
                         "CD200R3", # Mast cells
                         "HLA-DRA" # MHC-antigen presenting
                         ),
            order = T,
            ncol = 6)
```

# Getting data to evaluate integration
```{r}
# Save confounders metadata and PCA coordinates to evaluate integration
confounders_df <- seurat_obj@meta.data[, confounder_variables]
saveRDS(confounders_df, file = paste0(path_r_objects, "/CLARIAJOA_confounders.rds"))

# Get embeddings from non-integrated data
saveRDS(seurat_obj@reductions$pca@cell.embeddings[,1:20],
        file = paste0(path_r_objects, "/CLARIAJOA_pca_uncorrected.rds"))

# Get embeddings from integrated data
saveRDS(seurat_obj_integrated@reductions$harmony@cell.embeddings[,1:20],
        file = paste0(path_r_objects, "/CLARIAJOA_pca_integrated_harmony.rds"))
```

# Save processed Seurat objects
```{r}
# Save processed lognorm Seurat object
saveRDS(seurat_obj_integrated, paste0(path_r_objects, "/CLARIAJOA_harmony_processed.rds"))
#seurat_obj_integrated <- readRDS(paste0(path_r_objects, "/CLARIAJOA_harmony_processed.rds"))
```

# Session Info
```{r}
sessionInfo()
```
