---
title: "Annotation from lineages"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction

## Objective
In this Rmarkdown document, we are going to merge cell type identities annotated by independent cell-lineage analysis into a single one.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(DT)
library(grid)
library(RColorBrewer)
library(scales)
library(gtable)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/CLARIAJOA_metadata.csv")

path_r_objects <- here::here("04_clustering_annotation/results/R_objects")
path_r_figs <- here::here("04_clustering_annotation/results/figs")
path_r_tables <- here::here("04_clustering_annotation/results/tables")

# Functions
source(here::here("bin/utils.R"))
`%ni%` <- Negate(`%in%`)

confounder_variables <- c("library_name", "hashing", "patient", "treatment", "disease_grade", "gender")
confounder_names <- c("Library", "Hashing", "Patient", "Treatment", "Disease", "Gender")
qc_metrics <-  c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb", "S.Score", "G2M.Score")

# Color palette
color_palette <- Polychrome::createPalette(60, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```

## Load data
The data used in this Rmarkdown document comes from **CLARIAJOA** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)
metadata <- metadata %>% filter(type == "cDNA")
print("Libraries metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))
```


```{r}
# Load Seurat object -- ANNOT LEVEL 1
seurat_obj <- readRDS(paste0(path_r_objects, "/CLARIAJOA_clustering_level1_annotation.rds"))

# Diet Seurat object
seurat_obj <- DietSeurat(
  seurat_obj,
  counts = TRUE,
  data = TRUE,
  scale.data = FALSE,
  assays = "RNA",
  dimreducs = c("pca", "umap", "harmony"),
  graphs = c("RNA_nn", "RNA_snn"))

# Preserving only metadata of interest
seurat_metadata_variables <- c(names(seurat_obj@meta.data[1:38]), "annot_level1")
seurat_obj@meta.data <- seurat_obj@meta.data[seurat_metadata_variables]

seurat_obj
```


# Add cell-types identity

## Level 1
```{r}
table(seurat_obj$annot_level1)
```

## Level 2
```{r}
annotations_full <- data.frame(annot_level1 = character(),
                               annot_level2 = character()
)

level1_list <- c("Myeloid cells", "T-NK cells", "B Lymphocytes", "Mixed cells")
level1_filenames <- c("myeloidcells", "TNKcells", "Blymphocytes", "mixedcells")

# Load & add annotations & merge all together
for (i in 1:length(level1_list)) {
  annot1 <- level1_list[i]
  filename <- level1_filenames[i]
  file_path <- paste0(path_r_objects, "/CLARIAJOA_", filename, "_annotation_level2.rds")
  # loading annotations
  annot <- readRDS(file_path)
  # adding annot_level2 labels
  annotations_full <- rbind(annotations_full, annot)
}


# Add metadata to seurat object
annotations_full <- annotations_full["annot_level2"]
seurat_obj <- AddMetaData(seurat_obj, metadata = annotations_full)

# Replace NA's by level1
seurat_obj$annot_level2[is.na(seurat_obj$annot_level2)] <- seurat_obj$annot_level1[is.na(seurat_obj$annot_level2)]

table(seurat_obj$annot_level2)
length(unique(seurat_obj$annot_level2))
```


## Level 3
```{r}
annotations_full <- data.frame(annot_level1 = factor(),
                               annot_level2 = factor(),
                               annot_level3 = factor()
)

lineage_list <- list("Myeloid cells", "B Lymphocytes", "CD4+ T cells", "CD8+ T cells", "NK cells")
lineage_filenames <- c("myeloidcells", "Blymphocytes", "CD4Tcells", "CD8Tcells", "NKcells")

# Load & add annotations & merge all together
for (i in 1:length(lineage_list)) {
  lineage <- lineage_list[[i]]
  filename <- lineage_filenames[i]
  file_path <- paste0(path_r_objects, "/CLARIAJOA_", filename, "_annotation_level3.rds")
  # loading annotations
  annot <- readRDS(file_path)
  # adding annot_level3 labels
  annotations_full <- rbind(annotations_full, annot)
}

# Add metadata to seurat object
annotations_full <- annotations_full["annot_level3"]
seurat_obj <- AddMetaData(seurat_obj, metadata = annotations_full)

# Replace level3 NA by level2 values
#seurat_obj$annot_level3 <- as.factor(seurat_obj$annot_level3)
seurat_obj$annot_level3[is.na(seurat_obj$annot_level3)] <- seurat_obj$annot_level2[is.na(seurat_obj$annot_level3)]

table(seurat_obj$annot_level3)
length(unique(seurat_obj$annot_level3))
```


## Level 4
```{r}
annotations_full <- data.frame(annot_level1 = factor(),
                               annot_level2 = factor(),
                               annot_level3 = factor(),
                               annot_level3 = factor()
)

lineage_list <- list("CD4+ T cells", "CD8+ T cells", "NK cells")
lineage_filenames <- c("CD4Tcells", "CD8Tcells", "NKcells")

# Load & add annotations & merge all together
for (i in 1:length(lineage_list)) {
  lineage <- lineage_list[[i]]
  filename <- lineage_filenames[i]
  file_path <- paste0(path_r_objects, "/CLARIAJOA_", filename, "_annotation_level4.rds")
  # loading annotations
  annot <- readRDS(file_path)
  # adding annot_level3 labels
  annotations_full <- rbind(annotations_full, annot)
}

# Add metadata to seurat object
annotations_full <- annotations_full["annot_level4"]
seurat_obj <- AddMetaData(seurat_obj, metadata = annotations_full)

# Replace level3 NA by level2 values
#seurat_obj$annot_level3 <- as.factor(seurat_obj$annot_level3)
seurat_obj$annot_level4[is.na(seurat_obj$annot_level4)] <- seurat_obj$annot_level3[is.na(seurat_obj$annot_level4)]

table(seurat_obj$annot_level4)
length(unique(seurat_obj$annot_level4))
```

## CELL LINEAGES
```{r}
seurat_obj$cell_lineages <- as.character(seurat_obj$annot_level4)

seurat_obj$cell_lineages[seurat_obj$annot_level4 %in% c(
  "Naive B cells", "Activated Naive B cells", "Activated B cells", "Unswitched Memory B cells", "Switched Memory B cells", 
  "Transitional 1 B cells", "Transitional 2 B cells",
  "CD19hiCD11c+ Memory B cells", "Plasma cells", "Plasma cells cycling")] <- "B Lymphocytes"

seurat_obj$cell_lineages[seurat_obj$annot_level4 %in% c(
  "Naive CD4+ T cells", "Central Memory CD4+ T cells", "Central Memory ITGB1+ CD4+ T cells",
  "Effector Memory CD4+ T cells", "Activated Memory CD4+ T cells", "CD25-Foxp3+ Treg", "CD25+Foxp3+ Treg",
  "Th1", "Th17", "Proliferative CD4+ T cells")] <- "CD4+ T cells"

seurat_obj$cell_lineages[seurat_obj$annot_level4 %in% c(
  "Naive CD8+ T cells", "Activated CD8+ T cells", "Central Memory CD8+ T cells", "Transitional Memory CD8+ T cells",
  "Effector Memory CD8+ T cells", "Effector Activated CD8+ T cells", "Cytotoxic CD8+ T cells", "IFN+ CD8+ T cells",
  "Proliferative CD8+ T cells")] <- "CD8+ T cells"

seurat_obj$cell_lineages[seurat_obj$annot_level4 %in% c(
  "dN HAVCR2+ T cells", "gd T cells", "MAIT cells", "Innate-like T cells", "NK-like T cells")] <- "Unconventional T cells"

seurat_obj$cell_lineages[seurat_obj$annot_level4 %in% c(
  "CD56bright NK cells", "CD56dim CD16+CD57- NK cells", "CD56dim CD16+CD57-CXCR4+ NK cells",
  "CD56dim CD16+CD57+ NK cells","CD56dim CD16+CD2+ NK cells", "CD56dim CD16+IFNg+ NK cells",
  "NK cells cycling")] <- "NK cells"

seurat_obj$cell_lineages[seurat_obj$annot_level4 %in% c(
  "Classical Monocytes", "Intermediate Monocytes HAVCR2+", "Intermediate Monocytes Inflammatory",
  "Intermediate Monocytes IFN+", "Non-Classical Monocytes", "Macrophages-like Monocytes", "DC2",
  "DC4 FCGR3B+", "pDCs", "GMP cells", "Early-MDSC", "G-MDSC", "G-MDSC CST7+")] <- "Myeloid cells"

seurat_obj$cell_lineages[seurat_obj$annot_level4 %in% ("Granulocytes")] <- "Granulocytes"

seurat_obj$cell_lineages[seurat_obj$annot_level4 %in% c(
  "Platelet complexes", "Complex Platelets-B cells", "Complex Platelets-T cells", 
  "Complex Platelets-Classical Monocytes", "Complex Platelets-Non-Classical Monocytes", "Complex Platelets-DCs")] <- "Platelet complexes"

seurat_obj$cell_lineages <- as.factor(seurat_obj$cell_lineages)
table(seurat_obj$cell_lineages)
```

## CELL TYPES
```{r}
seurat_obj$cell_types <- as.character(seurat_obj$annot_level4)
seurat_obj$cell_types <- as.factor(seurat_obj$cell_types)
table(seurat_obj$cell_types)
```

## Remove annotation slots
```{r}
seurat_obj$annot_level1 <- NULL
seurat_obj$annot_level2 <- NULL
seurat_obj$annot_level3 <- NULL
seurat_obj$annot_level4 <- NULL
```


# Save data
```{r}
system_date <- gsub("-","",Sys.Date())
print(system_date)

system_date <- "20221102"
```

## Full dataset
```{r}
# Save metadata
saveRDS(seurat_obj@meta.data, paste0(path_r_objects, "/CLARIAJOA_metadata_full_", system_date, ".rds"))
openxlsx::write.xlsx(seurat_obj@meta.data, paste0(path_r_tables, "/CLARIAJOA_metadata_full_", system_date, ".xlsx"))

# Save annotated Seurat object
saveRDS(seurat_obj, paste0(path_r_objects, "/CLARIAJOA_clustering_annotation_full_", system_date, ".rds"))
#seurat_obj <- readRDS(paste0(path_r_objects, "/CLARIAJOA_clustering_annotation_full_20221102.rds"))
```

## Remove uninteresting cells
```{r}
Idents(seurat_obj) <- "cell_lineages"
cells_to_remove <- c("Doublets", "Low quality cells", "Platelet complexes")
seurat_obj_cleaned <- subset(seurat_obj, idents = cells_to_remove, invert = TRUE)

seurat_obj_cleaned$cell_lineages <- as.factor(as.character(seurat_obj_cleaned$cell_lineages))
seurat_obj_cleaned$cell_types <- as.factor(as.character(seurat_obj_cleaned$cell_types))

seurat_obj_cleaned
table(seurat_obj_cleaned$cell_types)
length(unique(seurat_obj_cleaned$cell_types))
rm(seurat_obj)
```

```{r}
# Reordering condition
seurat_obj_cleaned$disease_grade <- factor(x = seurat_obj_cleaned$disease_grade,
                                        levels = c(
                                        "Healthy",
                                        "AD",
                                        "ACLFg1",
                                        "ACLFg2",
                                        "ACLFg3"
                                        ))

seurat_obj_cleaned$disease_grade_treatment <- factor(x = seurat_obj_cleaned$disease_grade_treatment,
                                        levels = c(
                                        "Healthy_vehicle",
                                        "AD_vehicle",
                                        "ACLFg1_vehicle",
                                        "ACLFg2_vehicle",
                                        "ACLFg3_vehicle",
                                        "Healthy_HSA",
                                        "AD_HSA",
                                        "ACLFg1_HSA",
                                        "ACLFg2_HSA",
                                        "ACLFg3_HSA"
                                        ))

# Reordering lineages
seurat_obj_cleaned$cell_lineages <- factor(x = seurat_obj_cleaned$cell_lineages,
                                        levels = c(
                                        "B Lymphocytes",
                                        "Myeloid cells",
                                        "Granulocytes",
                                        "CD4+ T cells",
                                        "CD8+ T cells",
                                        "Unconventional T cells",
                                        "NK cells"
                                        ))

# Reordering cells
seurat_obj_cleaned$cell_types <- factor(x = seurat_obj_cleaned$cell_types,
                                        levels = c(
                                        # B Lymphocytes
                                        "Transitional 1 B cells",
                                        "Transitional 2 B cells",
                                        "Naive B cells",
                                        "Activated Naive B cells",
                                        "Activated B cells",
                                        "Unswitched Memory B cells",
                                        "Switched Memory B cells",
                                        "CD19hiCD11c+ Memory B cells",
                                        "Plasma cells",
                                        "Plasma cells cycling",
                                         # Myeloid cells
                                        "Classical Monocytes",
                                        "Intermediate Monocytes HAVCR2+",
                                        "Intermediate Monocytes Inflammatory",
                                        "Intermediate Monocytes IFN+",
                                        "Non-Classical Monocytes",
                                        "Macrophages-like Monocytes",
                                        "DC2",
                                        "DC4 FCGR3B+",
                                        "pDCs",
                                        "GMP cells",
                                        "Early-MDSC",
                                        "G-MDSC",
                                        "G-MDSC CST7+",
                                        "Granulocytes",
                                        # CD4+ T cells
                                        "Naive CD4+ T cells",
                                        "Central Memory CD4+ T cells",
                                        "Central Memory ITGB1+ CD4+ T cells",
                                        "Effector Memory CD4+ T cells",
                                        "Activated Memory CD4+ T cells",
                                        "CD25-Foxp3+ Treg",
                                        "CD25+Foxp3+ Treg",
                                        "Th1",
                                        "Th17",
                                        "Proliferative CD4+ T cells",
                                        # CD8+ T cells
                                        "Naive CD8+ T cells",
                                        "Activated CD8+ T cells",
                                        "Central Memory CD8+ T cells",
                                        "Transitional Memory CD8+ T cells",
                                        "Effector Memory CD8+ T cells",
                                        "Effector Activated CD8+ T cells",
                                        "Cytotoxic CD8+ T cells",
                                        "IFN+ CD8+ T cells",
                                        "Proliferative CD8+ T cells",
                                        # Unconventional T cells
                                        "dN HAVCR2+ T cells",
                                        "gd T cells",
                                        "MAIT cells",
                                        "Innate-like T cells",
                                        "NK-like T cells",
                                        # NK cells
                                        "CD56bright NK cells",
                                        "CD56dim CD16+CD57- NK cells",
                                        "CD56dim CD16+CD57-CXCR4+ NK cells",
                                        "CD56dim CD16+CD57+ NK cells", 
                                        "CD56dim CD16+CD2+ NK cells",
                                        "CD56dim CD16+IFNg+ NK cells",
                                        "NK cells cycling"
                                        # Platelets
                                        #"Platelet complexes",
                                        #"Complex Platelets-B cells",
                                        #"Complex Platelets-T cells",
                                        #"Complex Platelets-Classical Monocytes",
                                        #"Complex Platelets-Non-Classical Monocytes",
                                        #"Complex Platelets-DCs"
                                        ))
```

## Cleaned dataset
```{r}
# Save metadata
saveRDS(seurat_obj_cleaned@meta.data, paste0(path_r_objects, "/CLARIAJOA_metadata_cleaned_", system_date, ".rds"))
openxlsx::write.xlsx(seurat_obj_cleaned@meta.data, paste0(path_r_tables, "/CLARIAJOA_metadata_cleaned_", system_date, ".xlsx"))

# Save annotated Seurat object
saveRDS(seurat_obj_cleaned, paste0(path_r_objects, "/CLARIAJOA_clustering_annotation_cleaned_", system_date, ".rds"))
#seurat_obj_cleaned <- readRDS(paste0(path_r_objects, "/CLARIAJOA_clustering_annotation_cleaned_", system_date, ".rds"))
```


# Data visualization
```{r}
# Assign lineages colors
lineages_color_palette <- c("B Lymphocytes" = "#F94144",
                            "Myeloid cells" = "#90BE6D",
                            "Granulocytes" = "#132A13",
                            "CD4+ T cells" = "#277DA1",
                            "CD8+ T cells" = "#F9C74F",
                            "Unconventional T cells" = "#9F86C0",
                            "NK cells" = "#F9844A"
)

# Assign cell-types colors
cells_color_palette <- c( # B cells
                          "Transitional 1 B cells" = "#fa788b",
                          "Transitional 2 B cells" = "#641220",
                          "Naive B cells" = "#641220",
                          "Activated Naive B cells" = "#825a5f",
                          "Activated B cells" = "#E01E37",
                          "Unswitched Memory B cells" = "#59010f",
                          "Switched Memory B cells" = "#85182A",
                          "CD19hiCD11c+ Memory B cells" = "#A71E34",
                          "Plasma cells" = "#E5383B",
                          "Plasma cells cycling" = "#BF4342",
                          # Myeloid
                          "Classical Monocytes" = "#52B788",
                          "Intermediate Monocytes HAVCR2+" = "#1B4332",
                          "Intermediate Monocytes Inflammatory" = "#2D6A4F",
                          "Intermediate Monocytes IFN+" = "#40916C",
                          "Non-Classical Monocytes" = "#081C15",
                          "Macrophages-like Monocytes" = "#D8F3DC",
                          "DC2" = "#31572C",
                          "DC4 FCGR3B+" = "#4F772D",
                          "pDCs" = "#90A955",
                          "GMP cells" = "#3E5219",
                          "Early-MDSC" = "#2B9348",
                          "G-MDSC" = "#80B918",
                          "G-MDSC CST7+" = "#007F5F",
                          "Granulocytes" = "#132A13",
                          # CD4+ T cells
                          "Naive CD4+ T cells" = "#64B5F6",
                          "Central Memory CD4+ T cells" = "#1976D2",
                          "Central Memory ITGB1+ CD4+ T cells" = "#2196F3",
                          "Effector Memory CD4+ T cells" = "#0D47A1",
                          "Activated Memory CD4+ T cells" = "#BBDEFB",
                          "CD25-Foxp3+ Treg" = "#788BFF",
                          "CD25+Foxp3+ Treg" = "#5465FF",
                          "Th1" = "#E2FDFF",
                          "Th17" = "#BFD7FF",
                          "Proliferative CD4+ T cells" = "#9BB1FF",
                          # CD8+ T cells
                          "Naive CD8+ T cells" = "#583101",
                          "Activated CD8+ T cells" = "#A47148",
                          "Central Memory CD8+ T cells" = "#6F4518",
                          "Transitional Memory CD8+ T cells" = "#8B5E34",
                          "Effector Memory CD8+ T cells" = "#BC8A5F",
                          "Effector Activated CD8+ T cells" = "#D4A276",
                          "Cytotoxic CD8+ T cells" = "#C36F09",
                          "IFN+ CD8+ T cells" = "#F4E409",
                          "Proliferative CD8+ T cells" = "#EEBA0B",
                          # Unconventional T cells
                          "dN HAVCR2+ T cells" = "#FF4800",
                          "gd T cells" = "#FF6000",
                          "MAIT cells" = "#FF7900",
                          "Innate-like T cells" = "#FF9100",
                          "NK-like T cells" = "#FFAA00",
                          # NK cells
                          "CD56bright NK cells" = "#10002B",
                          "CD56dim CD16+CD57- NK cells" = "#240046",
                          "CD56dim CD16+CD57-CXCR4+ NK cells" = "#E0AAFF",
                          "CD56dim CD16+CD57+ NK cells" = "#7B2CBF", 
                          "CD56dim CD16+CD2+ NK cells" = "#9D4EDD",
                          "CD56dim CD16+IFNg+ NK cells" = "#C77DFF",
                          "NK cells cycling" = "#5A189A"
                          )
```

## UMAP colors
### Lineages
```{r fig.width=16, fig.height=12}
# Clustering overview
gg_umap_clustering_cellsubtypes_cleaned <- DimPlot(object = seurat_obj_cleaned,
        group.by = "cell_lineages",
        label = FALSE,
        cols = lineages_color_palette,
        raster = FALSE,
        shuffle = TRUE,
        pt.size = 1
        ) +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1)) +
  labs(title="Cell lineages - Cleaned")
gg_umap_clustering_cellsubtypes_cleaned

ggsave(filename = paste0(path_r_figs, "/CLARIAJOA_clustering_annotationlineages_cleaned_umap_", system_date, ".png"),
       plot = gg_umap_clustering_cellsubtypes_cleaned)
```


###  Cell-types
```{r fig.width=20, fig.height=12}
# Clustering overview
gg_umap_clustering_cellsubtypes_cleaned <- DimPlot(object = seurat_obj_cleaned,
        group.by = "cell_types",
        label = FALSE,
        cols = cells_color_palette,
        raster = FALSE,
        shuffle = TRUE,
        pt.size = 1
        ) +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 2)) +
  labs(title="Celltypes - Cleaned")
gg_umap_clustering_cellsubtypes_cleaned

ggsave(filename = paste0(path_r_figs, "/CLARIAJOA_clustering_annotationcells_cleaned_umap_", system_date, ".png"),
       plot = gg_umap_clustering_cellsubtypes_cleaned)
```

###  Lineages by Disease_grade
```{r fig.width=25, fig.height=5}
# Clustering overview
gg_umap_clustering_cellsubtypes_cleaned <- DimPlot(object = seurat_obj_cleaned,
        group.by = "cell_lineages",
        split.by = "disease_grade",
        label = FALSE,
        cols = lineages_color_palette,
        raster = FALSE,
        pt.size = 0.25,
        ncol = 5
        ) +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))
gg_umap_clustering_cellsubtypes_cleaned

ggsave(filename = paste0(path_r_figs, "/CLARIAJOA_split-DG_cleaned_umap_", system_date, ".png"),
       plot = gg_umap_clustering_cellsubtypes_cleaned)
```

```{r fig.width=25, fig.height=10}
# Clustering overview
gg_umap_clustering_cellsubtypes_cleaned <- DimPlot(object = seurat_obj_cleaned,
        group.by = "cell_lineages",
        split.by = "disease_grade_treatment",
        label = FALSE,
        cols = lineages_color_palette,
        raster = FALSE,
        pt.size = 0.25,
        ncol = 5
        ) +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))
gg_umap_clustering_cellsubtypes_cleaned

ggsave(filename = paste0(path_r_figs, "/CLARIAJOA_split-DGT_cleaned_umap_", system_date, ".png"),
       plot = gg_umap_clustering_cellsubtypes_cleaned)
```


# Celullar proportions
## Lineages
```{r fig.width=15, fig.height=6}
gg_cell_num <- dittoSeq::dittoBarPlot(seurat_obj_cleaned, 
                            "cell_lineages",
                            group.by = "disease_grade_treatment",
                            scale = "count",
                            legend.show = FALSE,
                            retain.factor.levels = TRUE,
                            color.panel = lineages_color_palette)

gg_cell_prop <- dittoSeq::dittoBarPlot(seurat_obj_cleaned, 
                             "cell_lineages",
                             group.by = "disease_grade_treatment",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = lineages_color_palette)

gg_cell_num + gg_cell_prop
```

## Cell-types
```{r fig.width=15, fig.height=6}
dittoSeq::dittoBarPlot(seurat_obj_cleaned, 
                            "cell_types",
                            group.by = "disease_grade_treatment",
                            scale = "count",
                            #legend.show = FALSE,
                            retain.factor.levels = TRUE,
                            color.panel = cells_color_palette)

dittoSeq::dittoBarPlot(seurat_obj_cleaned, 
                             "cell_types",
                             group.by = "disease_grade_treatment",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = cells_color_palette)
```

# Gene markers plot
```{r}
# Gene markers for clusters
# TOP 5 (for subsets) --> reduced to TOP 2 (for whole dataset)
lineages_gene_markers <- c(
  "CD79A", "CD79B",	"MS4A1", "JCHAIN", ## B Lymphocytes
  "CD14",	"LYZ", "FCER1A", "FCGR3B", ## Myeloid cells
  "GATA2",	"RGS2", # Granulocytes
  "CD3E",	"CD3D",	"CD4", ## CD4+ CELLS	
  "CD8A", "CD8B", ## CD8+ CELLS		
  "SELL", "PASK","KLRB1",
  "NCAM1",	"FCGR3A", "KLRB1" ## NK cells		
)

cells_gene_markers <- c(
  "CD79A", "CD79B",	"CD19",	"MS4A1", ## B Lymphocytes
  "TCL1A",	"VPREB3",	"LTB",	"FCRL5",
  "IGHD",	"IGHM",	"SELL",	"CCR7",
  "CD69",	"CD83",	"MYC",	"HLA-DQA1",
  "CD27", "S100A10",	"SSPN",
  "ITGAX",	"HLA-DPA1", "CD38",
  "JCHAIN", "IGHG1",	"XBP1",

  "CD14",	"LYZ"	,"FCGR3A", ## Myeloid cells
  "VCAN",	"EREG",
  "HAVCR2",	"HLA-DRA",	"HLA-DPA1",
  "CCL3",	"IL1B",	"TNF",	"CXCL2",
  "ISG15",	"IFIT2",	"IFIT3",	"STAT1",
  "MS4A7",	"LST1",
  "C1QA",	"SELENOP",	"MRC1",	"CD163",
  "FCER1A",	"IL7R",	"CLEC10A",
  "FCGR3B",	"CSF3R",	"IFITM2",
  "JCHAIN",	"IL3RA",	"IRF8",
  "DEFA3",	"DEFA4",	"MS4A3",	"RETN",
  "MMP9",	"LCN2",	"S100A12",
  "CST7",
  "GATA2",	"RGS2", # Granulocytes
  
  "CD3E",	"CD3D",	"CD4", ## CD4+ CELLS	
  "SELL",	"CCR7",	"LEF1",	"TCF7",
  "IL7R",	"PASK",		
  "ITGB1",	"LGALS1",	"ANXA1",	
  "KLRB1",	"MAF",	"S100A4", "IL32",
  "CD69",	"DUSP4",	"CREM",	
  "FOXP3",	"DUSP",	"CTLA4",	
  "IL2RA",	"BATF",	"TIGIT",	
  "CCL5",	"CXCR5",	"TBX21",	
  "CCR6",	"RORC",		

  "CD8A", "CD8B", ## CD8+ CELLS		
  "SELL",	"CCR7",	"LEF1",	"TCF7",
  "CD69",	"DUSP4", "CREM",
  "IL7R",	"PASK",		
  "ZEB2",			
  "KLRB1",	"MAF",	"S100A4",	"IL32",
  "DUSP2",	"RGS1",	"GZMK",	
  "ISG15",	"IFIT2",	"IFIT3",	"STAT1",
  "GLYN",	"CCL5",	"NKG7",	
  
  "HAVCR2",	"GZMK",	"ACTG1",	
  "TRDC",	"TRGC1",	"KLRC1",	
  "KLRB1",	"IL7R",	"CCR6",	"GZMK",
  "FCER1G",	"TCR",	"TSHZ2",
  "FCGR3A",	"GNLY",	"KLRD1",	
  
  "NCAM1",	"FCGR3A", "KLRB1", ## NK cells		
  "GPR183",	"IL7R",	"XCL1",	"XCL2",
  "GZMK",	"VIM",	"MT2A",
  "CXCR4",	"NFKBIA",		
  "B3GAT1",	"CX3CR1",		
  "CD2",
  "IFNG",	"CCL3",	"CCL4",	

  "STMN1",	"TOP2A",	"MKI67"	
)

lineages_gene_markers <- lineages_gene_markers[!duplicated(lineages_gene_markers)]
cells_gene_markers <- cells_gene_markers[!duplicated(cells_gene_markers)]
```


### Lineages

```{r fig.width=12, fig.height=6}
dotplot_lineages <- DotPlot(seurat_obj_cleaned,
        features = lineages_gene_markers,
        group.by = "cell_lineages",
        #scale = TRUE,
        #col.max = 3,
        #col.min = 0,
        cols = "RdYlBu",
        dot.scale = 10) +
  #coord_flip() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5))

dotplot_lineages
```

```{r fig.width=10, fig.height=15}
avgexpr_mat <- AverageExpression(
  features = lineages_gene_markers,
  seurat_obj_cleaned,
  assays = "RNA",
  return.seurat = F,
  group.by = "cell_lineages",
  slot = "counts")

cell_types <- names(table(seurat_obj_cleaned$cell_lineages))
mycolors <- list(cell_type = lineages_color_palette)
mat <- as.data.frame(avgexpr_mat$RNA)
colnames(mat) <- cell_types
annotation_col = data.frame(
                    cell_type = cell_types) 
rownames(annotation_col) <- cell_types

input_mat <- t(apply(mat, MARGIN = 1, FUN = function(X) (X - min(X))/diff(range(X))))
```


```{r fig.width=4, fig.height=8}
#pdf(paste(figures_folder, "M6B_reduced_genes_heatmap.pdf", sep=""),width=5,height=7,paper='special') 
pheatmap2::pheatmap2(input_mat,
                annotation_col = annotation_col,
                annotation_colors = mycolors,
                annotation_names_col = F,
                annotation_legend = F,
                show_rownames=T, show_colnames = F, 
                border_color = NA,
                cluster_rows = F,
                cluster_cols = F,
                fontsize_row = 6,
                gaps_row = c(4,8,10,17), 
                gaps_col = c(1,3,6) 
)

grid::grid.ls(grid.force()) # "col_annotation" looks like it's the one to edit
#grid::grid.gedit("annotation_col", gp = gpar(col="black"))
#dev.off()
```



### Cell-types

```{r fig.width=25, fig.height=12}
dotplot_cells <- DotPlot(seurat_obj_cleaned,
        features = cells_gene_markers,
        group.by = "cell_types",
        #scale = TRUE,
        #col.max = 3,
        #col.min = 0,
        cols = "RdYlBu",
        dot.scale = 5) +
  #coord_flip() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5))

dotplot_cells
```


```{r fig.width=10, fig.height=15}
avgexpr_mat <- AverageExpression(
  features = cells_gene_markers,
  seurat_obj_cleaned,
  assays = "RNA",
  return.seurat = F,
  group.by = "cell_types",
  slot = "counts")

cell_types <- names(table(seurat_obj_cleaned$cell_types))
mycolors <- list(cell_type = cells_color_palette)
mat <- as.data.frame(avgexpr_mat$RNA)
colnames(mat) <- cell_types
annotation_col = data.frame(
                    cell_type = cell_types) 
rownames(annotation_col) <- cell_types

input_mat <- t(apply(mat, MARGIN = 1, FUN = function(X) (X - min(X))/diff(range(X))))
```


```{r fig.width=7, fig.height=12}
#pdf(paste(figures_folder, "M6B_reduced_genes_heatmap.pdf", sep=""),width=5,height=7,paper='special') 
pheatmap2::pheatmap2(input_mat,
                annotation_col = annotation_col,
                annotation_colors = mycolors,
                annotation_names_col = F,
                annotation_legend = F,
                show_rownames=T, show_colnames = F, 
                border_color = NA,
                cluster_rows = F,
                cluster_cols = F,
                fontsize_row = 6,
                gaps_row = c(25,64,89,95,104), 
                gaps_col = c(7,21,31,38,45) 
)

grid::grid.ls(grid.force()) # "col_annotation" looks like it's the one to edit
#grid::grid.gedit("annotation_col", gp = gpar(col="black"))
#dev.off()
```

# Session Info
```{r}
sessionInfo()
```
